### 2025-07-01 machine-learning.qmd
20250630 - free up memory

@@

```{r}
# Prepare data for merging
#-todo: calculate years_of_experience
## Use common name for the same (gsis_id) ID variable

@@

@@

```{r}
# Create lists of objects to merge, depending on data structure: id; or id-season; or id-season-week
#-todo: remove redundant variables
playerListToMerge <- list(
  nfl_players %>% filter(!is.na(gsis_id)),
  nfl_draftPicks %>% filter(!is.na(gsis_id)) %>% select(-season)
)

@@

@@

```{r}
seasonalData <- powerjoin::power_full_join(
  playerSeasonMerged,
  playerMerged %>% select(-age, -years_of_experience, -team, -team_abbr, -team_seq, -current_team_id), # drop variables from id objects that change from year to year (and thus are not necessarily accurate for a given season)
  by = "gsis_id",
  conflict = coalesce_xy # where the objects have the same variable name (e.g., position), keep the values from object 1, unless it's NA, in which case use the relevant value from object 2
) %>% 
  filter(!is.na(season)) %>% 
  select(gsis_id, season, player_display_name, position, team, games, everything())
```

@@

```{r}
seasonalAndWeeklyData <- powerjoin::power_full_join(
  playerSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy # where the objects have the same variable name (e.g., position), keep the values from object 1, unless it's NA, in which case use the relevant value from object 2
) %>% 
  filter(!is.na(week)) %>% 
  select(gsis_id, season, week, full_name, position, team, everything())
```

@@

```{r}
# Convert character and logical variables to factors
seasonalData <- seasonalData %>% 
  mutate(
    across(
      where(is.character),
      as.factor
    ),
    across(
      where(is.logical),
      as.factor
    )
  )
```

@@

```{r}
seasonalData <- seasonalData %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  fill(
    player_name, player_display_name, pos, position, position_group,
    .direction = "downup") %>% 
  ungroup()
```

@@

```{r}
seasonalData_lag <- seasonalData %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  mutate(
    fantasyPoints_lag = lead(fantasyPoints)
  ) %>% 
  ungroup()

@@

@@

```{r}
seasonalData_lag %>% select_if(~class(.) == "Date")
seasonalData_lag %>% select_if(is.character)
seasonalData_lag %>% select_if(is.factor)
seasonalData_lag %>% select_if(is.logical)

@@

@@

```{r}
seasonalData_lag_subsetQB <- seasonalData_lag_subset %>% 
  filter(position == "QB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints, fantasyPoints_lag,
    completions:rushing_2pt_conversions, special_teams_tds, contains(".pass"), contains(".rush"))

@@

@@

```{r}
seasonalData_lag_qb_all <- seasonalData_lag_subsetQB
seasonalData_lag_rb_all <- seasonalData_lag_subsetRB
seasonalData_lag_wr_all <- seasonalData_lag_subsetWR
seasonalData_lag_te_all <- seasonalData_lag_subsetTE

@@

@@

```{r}
# QBs
seasonalData_lag_qb_all_imp <- missRanger::missRanger(
  seasonalData_lag_qb_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
# RBs
seasonalData_lag_rb_all_imp <- missRanger::missRanger(
  seasonalData_lag_rb_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
# WRs
seasonalData_lag_wr_all_imp <- missRanger::missRanger(
  seasonalData_lag_wr_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
# TEs
seasonalData_lag_te_all_imp <- missRanger::missRanger(
  seasonalData_lag_te_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

@@

```{r}
#| include: false

@@

```{r}
cl <- parallel::makeCluster(num_cores)
doParallel::registerDoParallel(cl)

@@

```{r}
pred_vars_qb <- data_train_qb_matrix %>% 
  as_tibble() %>% 
  select(-fantasyPoints_lag, -ageCentered20, ageCentered20Quadratic) %>% # -gsis_id
  names()

@@

```{r}
model_likelihood <- "gaussian" # gamma
nrounds <- 2000
```

@@

```{r}
# Partition training data into inner training data and validation data
ntrain_qb <- dim(data_train_qb_matrix)[1]

@@

@@

```{r}
gp_model_qb <- gpboost::GPModel(
  group_data = data_train_qb_matrix[,"gsis_id"],
  likelihood = model_likelihood,
  group_rand_coef_data = cbind(
    data_train_qb_matrix[,"ageCentered20"],
    data_train_qb_matrix[,"ageCentered20Quadratic"]),
  ind_effect_group_rand_coef = c(1,1))

@@

```{r}
gp_model_fit_qb <- gpboost::gpb.train(
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds_qb,
  params = params_qb) # verbose = 0

@@

@@

```{r}
# Test Model on Test Data
pred_test_qb <- predict(
  gp_model_fit_qb,
  data = data_test_qb_matrix[,pred_vars_qb],
  group_data_pred = data_test_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    data_test_qb_matrix[,"ageCentered20"],
    data_test_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

@@

```{r}
# Generate model predictions for next season
pred_nextYear_qb <- predict(
  gp_model_fit_qb,
  data = newData_qb_matrix[,pred_vars_qb],
  group_data_pred = newData_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    newData_qb_matrix[,"ageCentered20"],
    newData_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

>>>>

```{r}
#| label: free-up-memory1-ml
#| include: false

gc()
```

@@

```{r}
#| label: prepare-data-merging-machine-learning

# Prepare data for merging
#-todo: calculate years_of_experience
## Use common name for the same (gsis_id) ID variable

@@

```{r}
#| label: free-up-memory2-ml
#| include: false

gc()
```

@@

```{r}
#| label: merge-data1-machine-learning

# Create lists of objects to merge, depending on data structure: id; or id-season; or id-season-week
#-todo: remove redundant variables
playerListToMerge <- list(
  nfl_players %>% filter(!is.na(gsis_id)),
  nfl_draftPicks %>% filter(!is.na(gsis_id)) %>% select(-season)
)

@@

```{r}
#| label: free-up-memory3-ml
#| include: false

gc()
```

@@

```{r}
#| label: merge-data2-machine-learning

seasonalData <- powerjoin::power_full_join(
  playerSeasonMerged,
  playerMerged %>% select(-age, -years_of_experience, -team, -team_abbr, -team_seq, -current_team_id), # drop variables from id objects that change from year to year (and thus are not necessarily accurate for a given season)
  by = "gsis_id",
  conflict = coalesce_xy # where the objects have the same variable name (e.g., position), keep the values from object 1, unless it's NA, in which case use the relevant value from object 2
) %>% 
  filter(!is.na(season)) %>% 
  select(gsis_id, season, player_display_name, position, team, games, everything())
```

```{r}
#| label: free-up-memory4-ml
#| include: false

gc()
```

@@

```{r}
#| label: merge-data3-machine-learning

seasonalAndWeeklyData <- powerjoin::power_full_join(
  playerSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy # where the objects have the same variable name (e.g., position), keep the values from object 1, unless it's NA, in which case use the relevant value from object 2
) %>% 
  filter(!is.na(week)) %>% 
  select(gsis_id, season, week, full_name, position, team, everything())
```

```{r}
#| label: free-up-memory5-ml
#| include: false

gc()
```

@@

```{r}
#| label: additional-processing-machine-learning

# Convert character and logical variables to factors
seasonalData <- seasonalData %>% 
  mutate(
    across(
      where(is.character),
      as.factor
    ),
    across(
      where(is.logical),
      as.factor
    )
  )
```

```{r}
#| label: free-up-memory6-ml
#| include: false

gc()
```

@@

```{r}
#| label: fill-in-missing-data-machine-learning

seasonalData <- seasonalData %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  fill(
    player_name, player_display_name, pos, position, position_group,
    .direction = "downup") %>% 
  ungroup()
```

```{r}
#| label: free-up-memory7-ml
#| include: false

gc()
```

@@

```{r}
#| label: lag-fantasy-points-machine-learning

seasonalData_lag <- seasonalData %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  mutate(
    fantasyPoints_lag = lead(fantasyPoints)
  ) %>% 
  ungroup()

@@

```{r}
#| label: free-up-memory8-ml
#| include: false

gc()
```

@@

```{r}
#| label: subset-data-machine-learning

seasonalData_lag %>% select_if(~class(.) == "Date")
seasonalData_lag %>% select_if(is.character)
seasonalData_lag %>% select_if(is.factor)
seasonalData_lag %>% select_if(is.logical)

@@

```{r}
#| label: free-up-memory9-ml
#| include: false

gc()
```

@@

```{r}
#| label: separate-by-position-machine-learning

seasonalData_lag_subsetQB <- seasonalData_lag_subset %>% 
  filter(position == "QB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints, fantasyPoints_lag,
    completions:rushing_2pt_conversions, special_teams_tds, contains(".pass"), contains(".rush"))

@@

```{r}
#| label: free-up-memory10-ml
#| include: false

gc()
```

@@

```{r}
#| label: split-test-training-machine-learning

seasonalData_lag_qb_all <- seasonalData_lag_subsetQB
seasonalData_lag_rb_all <- seasonalData_lag_subsetRB
seasonalData_lag_wr_all <- seasonalData_lag_subsetWR
seasonalData_lag_te_all <- seasonalData_lag_subsetTE

@@

```{r}
#| label: free-up-memory11-ml
#| include: false

rm(nfl_players)
rm(nfl_teams)
rm(nfl_rosters)
rm(nfl_rosters_weekly)
rm(nfl_schedules)
rm(nfl_combine)
rm(nfl_draftPicks)
rm(nfl_depthCharts)
rm(nfl_pbp)
rm(nfl_4thdown)
rm(nfl_participation)
rm(nfl_actualFantasyPoints_weekly)
rm(nfl_injuries)
rm(nfl_snapCounts)
rm(nfl_espnQBR_seasonal)
rm(nfl_espnQBR_weekly)
rm(nfl_nextGenStats_weekly)
rm(nfl_advancedStatsPFR_seasonal)
rm(nfl_advancedStatsPFR_weekly)
rm(nfl_playerContracts)
rm(nfl_ftnCharting)
rm(nfl_playerIDs)
rm(nfl_rankings_draft)
rm(nfl_rankings_weekly)
rm(nfl_expectedFantasyPoints_weekly)
rm(nfl_expectedFantasyPoints_pbp)
rm(nfl_actualStats_career)
rm(nfl_actualStats_career_player)
rm(nfl_actualStats_career_player_inclPost)
rm(nfl_actualStats_seasonal)
rm(nfl_actualStats_seasonal_player)
rm(nfl_actualStats_seasonal_player_inclPost)
rm(nfl_actualStats_seasonal_team)
rm(nfl_actualStats_seasonal_team_inclPost)
rm(player_stats_weekly)
rm(player_stats_weekly_offense)
rm(player_stats_seasonal)
rm(player_stats_seasonal_offense)

rm(playerListToMerge)
rm(playerMerged)
rm(playerSeasonListToMerge)
rm(playerSeasonMerged)
rm(playerSeasonWeekListToMerge)
rm(playerSeasonWeekMerged)
rm(playerSeasonWeekPositionListToMerge)
rm(seasonalAndWeeklyData)

gc()
```

@@

```{r}
#| label: impute-qb-machine-learning

# QBs
seasonalData_lag_qb_all_imp <- missRanger::missRanger(
  seasonalData_lag_qb_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
#| label: free-up-memory12-ml
#| include: false

gc()
```

```{r}
#| label: impute-rb-machine-learning

# RBs
seasonalData_lag_rb_all_imp <- missRanger::missRanger(
  seasonalData_lag_rb_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
#| label: free-up-memory13-ml
#| include: false

gc()
```

```{r}
#| label: impute-wr-machine-learning

# WRs
seasonalData_lag_wr_all_imp <- missRanger::missRanger(
  seasonalData_lag_wr_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
#| label: free-up-memory14-ml
#| include: false

gc()
```

```{r}
#| label: impute-te-machine-learning

# TEs
seasonalData_lag_te_all_imp <- missRanger::missRanger(
  seasonalData_lag_te_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

@@

```{r}
#| label: free-up-memory15-ml
#| include: false

gc()
```

@@

```{r}
#| label: parallel-cores-machine-learning
#| include: false

@@

```{r}
#| label: cross-sectional-machine-learning

cl <- parallel::makeCluster(num_cores)
doParallel::registerDoParallel(cl)

@@

```{r}
#| label: specify-predictors-machine-learning

pred_vars_qb <- data_train_qb_matrix %>% 
  as_tibble() %>% 
  select(-fantasyPoints_lag, -ageCentered20, ageCentered20Quadratic) %>% # -gsis_id
  names()

@@

```{r}
#| label: specify-general-model-options-machine-learning

model_likelihood <- "gaussian" # gamma
nrounds <- 2000
```

@@

```{r}
#| label: tuning-parameters-machine-learning

# Partition training data into inner training data and validation data
ntrain_qb <- dim(data_train_qb_matrix)[1]

@@

```{r}
#| label: free-up-memory16-ml
#| include: false

gc()
```

@@

```{r}
#| label: model-parameters-machine-learning

@@

```{r}
#| label: fit-model-machine-learning

gp_model_fit_qb <- gpboost::gpb.train(
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds_qb,
  params = params_qb) # verbose = 0

@@

```{r}
#| label: free-up-memory17-ml
#| include: false

gc()
```

@@

```{r}
#| label: evaluate-model-machine-learning

# Test Model on Test Data
pred_test_qb <- predict(
  gp_model_fit_qb,
  data = data_test_qb_matrix[,pred_vars_qb],
  group_data_pred = data_test_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    data_test_qb_matrix[,"ageCentered20"],
    data_test_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

@@

```{r}
#| label: next-season-predictions-machine-learning

# Generate model predictions for next season
pred_nextYear_qb <- predict(
  gp_model_fit_qb,
  data = newData_qb_matrix[,pred_vars_qb],
  group_data_pred = newData_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    newData_qb_matrix[,"ageCentered20"],
    newData_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)
